---
title: "Project 2. The Infuence of Religiosity Dimensions on the Participation in Local Elections in the USA"
author: "Nevolin Dmutry"
output:
  md_document:
    variant: markdown_github
date: "2023-03-25"
---

This small research is based on the joint WVS-EVS dataset:
https://europeanvaluesstudy.eu/methodology-data-documentation/survey-2017/joint-evs-wvs/

```{r, message=FALSE, warning=FALSE, results='hide'}
library(haven)
library(tidyverse)
library(sjmisc)
library(car)
library(yardstick)
library(pROC)
library(effects)
library(sjPlot)
library(rpart)
library(rpart.plot)
library(polycor)
library(psych)
library(GPArotation)

usa1 <- read.csv("USA.csv")                                                                                 
```


In modern political science discourse, electoral behavior is a popular research object. Therefore, there are a lot of theoretical explanations, which factors mainly influence this behavior. One of the most supported theory, in my opinion, is based on the religiosity.

The relationship between religion and political participation has been rigorously investigated, typically employing basic measures of church attendance or denomination.It was done by many scientists, including R. Driskell, E. Embry and L. Lyon. They state that collective aspects of religious belief and practice often track with greater levels of political participation (Driskell, R., Embry, E., & Lyon, L., 2008).
Moreover, it is found that effect of religious practices can differ based on the specifics of a particular denomination. For example, L. E. Smith and L. D. Walker state that observed effect is really weak for Catholics (Smith, L. E., & Walker, L. D., 2013).

In this research I would like to check, how the level of religiosity, including both participation and value dimension, affects participation in local elections among US citizens. I find it extremely interesting mainly because of multicultural composition of the country (in case of religiosity), which, however, is dominated by protestant denomination. I look forward to really interesting finding, how, first of all, value dimension works in respect to local elections in the USA.

Therefore, my research question is following: **how religious belief and practice is associated with electoral participation in the USA**?

For analysing electoral participation, I work with the variable from the above mentioned data set, which contains respondents` answers on the question 
"When elections (local level) take place, do you vote always, usually or never?".

As predictors, I will use some socio-demographic variable as control ones, and variable related to religiosity.
Firstly, there will be some indicators, which describe religious participation:
- f028 - how often respondent attend religious services (apart from  weddings and funerals);
- f066_evs5 - how often respondent prays (apart from  weddings and funerals).

- f034 - does respondent find him- or herself religious or not;
- f063 - how God is important for respondent.

These are basic variables that are widely used in many researches dedicated to this topic. However, I decide to include some additions indicators, which represent belief dimension of religiosity:
- f053 - does respondent believe in hell or not;
- f054 - does respondent believe in heaven or not.
I guess that these beliefs in something after life can play controversial role in case of electoral participation:
from one hand, it can give additional impulses to participate in elections, however, it can also avoid people from participation in everything that is close to politics.
There is a threat that I will find big multicollinearity between these variables, but it worth to try to check their effects.

Based on selected indicators, I hypothesize that:
1. High level of religious participation is associated with the high level of electoral participation at the local level.
2. Strong religious beliefs are also associated with the high level of electoral participation at the local level.


```{r}
usa <- usa1 %>% filter(cntry_an == "US") %>%
  select(x001, x003, e263, f034, f028, f063, f066_evs5, f053, f054, f028b_wvs7) #choosing variables for US
```



Recoding elections variable:


First of all, this variable contains negative values that should be recoded as NAs:



```{r}
usa$elections <- ifelse(usa$e263 < 0, NA, usa$e263)
length(usa$elections) #153716
sum(is.na(usa$elections)) #5310
```

For now, this variable contains only about 33 missing values. However, I also want to recode as missings one category as a whole - people, who are not allowed to vote. Based on them, it is impossible to somehow evaluate the effect of religiosity.


```{r}
usa %>% filter(elections == 4) %>% select(elections) %>% count()
```


There are only about 88 of respondents who will be also excluded from the sample. It will not bring any bias in further analysis.


```{r}
usa$elections1 <- ifelse(usa$elections == 4, NA, usa$elections)
sum(is.na(usa$elections1)) #121

```

For now, the variable contains only about 121 missing values, which is a really good result after all preprocessing steps.


```{r}
descr(usa$elections1)
table(usa$elections1)
```


The distribution of values in this variable is a bit right-skewed: the minority of respondents belongs to the third category (they do not participate in local elections). In order to balance the distribution, I recode the variable into binary one. I divide respondents into two groups: those who always participate in local elections, and those participate not always or do not participate at all:


```{r}
usa$loc_elect <- ifelse(usa$elections1 > 1, 0, 1)

table(usa$elections1, usa$loc_elect)
  
```


Variable was recoded correctly.


```{r, message=FALSE, warning=FALSE}
usa %>%
  ggplot(aes(loc_elect)) +
  geom_bar(fill = 4, colour = 4) +
  theme_bw() +
  ylab("") +
  scale_x_continuous(name = "\nElectoral Participation\n", breaks = c(0, 1),
                     labels = c("Non-participation", "Participation"))+
  theme(axis.text.x = element_text(face = "bold", size = 10))
```



Values are distributed not equally, but, in general, sample is not extremely unbalanced.



As for other variables:



```{r}
unique(usa$x001)

names(usa)[names(usa) == "x001"] <- "sex" #gender variable
usa$sex <- ifelse(usa$sex == 2, "Female", "Male")
sum(is.na(usa$sex)) #0

usa %>%
  ggplot(aes(sex)) +
  geom_bar(fill = 4, colour = 4) +
  theme_bw() +
  ylab("")

```



In case of gender, the sample is balanced enough.



```{r, message=FALSE, warning=FALSE}
unique(usa$x003)

names(usa)[names(usa) == "x003"] <- "age" #age variable
sum(is.na(usa$age)) #0

usa %>%
  ggplot(aes(age)) +
  geom_histogram(bins = 15, fill = "white", colour = 4) +
labs(main = "",
         x = "\nAge\n") +
  theme_bw()

```



The distribution is a bit right-skewed, but it is okay for further analysis.



```{r}
names(usa)[names(usa) == "f034"] <- "relig" #is a respondent religious or not

unique(usa$relig) #it has missing values that are recoded as negative values

usa$relig <- ifelse(usa$relig < 0, NA, usa$relig) #negative values into NAs

usa$relig <- ifelse(usa$relig == 1, "Religious",
                       ifelse(usa$relig == 2, "Non-religious", "Atheist")) #labels instead of numbers

usa$relig <- as.factor(usa$relig) #as factor

sum(is.na(usa$relig)) #20, which is a small amount

usa %>%
  ggplot(aes(relig)) +
  geom_bar(fill = 4, colour = 4) +
  theme_bw() +
  labs(main = "",
       y = "",
       x = "")
  
```


First of all, this variable contains not so many missing values. Also, the distribution of this indicator is interesting: the majority of US respondents identify themselves as atheists and non-religious. It will be interesting to check, how the fact that respondent is religious according to his or her opinion affects electoral participation.

As for the "Service attendance" variable:


```{r}
names(usa)[names(usa) == "f028"] <- "services_attend" #how often respondent attends religious services

unique(usa$services_attend) #it has missing values that are recoded as negative values

usa$services_attend <- ifelse(usa$services_attend < 0, NA, usa$services_attend) #negative values into NAs

```



It is better to recode it as factor with labels instead of numbers, because it contains not so many unique levels (8) and intervals between them are not really equal to each other:



```{r}
usa$attend_services <- as.factor(ifelse(usa$services_attend == 1, "More than once a week",
                                 ifelse(usa$services_attend == 2, "Once a week",
                                        ifelse(usa$services_attend == 3, "Once a month",
                                               ifelse(usa$services_attend == 4, "Christmas or Easter day",
                                                      ifelse(usa$services_attend == 5, "Other holy days",
                                                              ifelse(usa$services_attend == 6, "Once a year",
                                                                     ifelse(usa$services_attend == 7, "Less often than once a year", "Practically never"))))))))

table(usa$attend_services, usa$services_attend)
```



Variable was recoded correctly.



```{r}
as.data.frame(table(usa$attend_services))
```


The majority of people attend such service practically never. However, other categories still contain enough observation for further analysis.



```{r, message=FALSE, warning=FALSE}

names(usa)[names(usa) == "f063"] <- "god_imp" #the importance of God

unique(usa$god_imp) #it has missing values that are recoded as negative values

usa$god_imp <- ifelse(usa$god_imp < 0, NA, usa$god_imp) #negative values into NAs

sum(is.na(usa$god_imp)) #18

usa %>%
  ggplot(aes(god_imp)) +
  geom_histogram(bins = 6, fill = "white", colour = 4) +
  labs(main = "",
         x = "\nImportans of God\n") +
  theme_bw()

```


Distribution of "God importance" variable does not look like a normal-one: the majority of values are in poles. However, it is not a problem for further analysis.


```{r}
names(usa)[names(usa) == "f028b_wvs7"] <- "pray" #how often respondent prays

unique(usa$pray) #it has missing values that are recoded as negative values

usa$pray <- ifelse(usa$pray < 0, NA, usa$pray) #negative values into NAs

sum(is.na(usa$pray)) #18

```



With "pray" variable we have the same situation as with religious services attendance. It is better to recode it as factor with labels instead of numbers, because it contains not so many unique levels (8) and intervals between them are not really equal to each other:



```{r}
usa$pray_oft <- as.factor(ifelse(usa$pray == 1, "More than once a week",
                                 ifelse(usa$pray == 2, "Once a week",
                                        ifelse(usa$pray == 3, "Once a month",
                                               ifelse(usa$pray == 4, "Christmas or Easter day",
                                                      ifelse(usa$pray == 5, "Other holy days",
                                                              ifelse(usa$pray == 6, "Once a year",
                                                                     ifelse(usa$pray == 7, "Less often than once a year", "Practically never"))))))))

table(usa$pray_oft, usa$pray)
```



Variable was recoded correctly.


```{r}
as.data.frame(table(usa$pray_oft))
```



Some categories here have not so many observations (like once a year, other holy days). However, it is still enough for further analysis.



```{r, message=FALSE, warning=FALSE}

names(usa)[names(usa) == "f053"] <- "hell" #does respondent believe in hell or not

unique(usa$hell) #it has missing values that are recoded as negative values

usa$hell <- ifelse(usa$hell < 0, NA, usa$hell) #negative values into NAs

usa$hell <- as.factor(ifelse(usa$hell == 0, "No", "Yes"))

usa$hell <- as.factor(usa$hell)

usa %>%
  ggplot(aes(hell)) +
  geom_bar(fill = 4, colour = 4) +
  theme_bw() +
  labs(main = "",
       y = "",
       x = "Believe in Hell")

```


This variable contains a small amount of missing values.
Interesting that the majority of respondents believe in hell according to their answers. In general, the variable is balanced enough for further analysis.


```{r, warning=FALSE, message=FALSE}

names(usa)[names(usa) == "f054"] <- "heaven" #does respondent believe in heaven or not

unique(usa$heaven) #it has missing values that are recoded as negative values

usa$heaven <- ifelse(usa$heaven < 0, NA, usa$heaven) #negative values into NAs

usa$heaven <- as.factor(ifelse(usa$heaven == 0, "No", "Yes"))

usa$heaven <- as.factor(usa$heaven)

usa %>%
  ggplot(aes(heaven)) +
  geom_bar(fill = 4, colour = 4) +
  theme_bw() +
  labs(main ="",
       y = "",
       x = "\nBelieve in Heaven\n")
```


This variable contains a small amount of missing values.
Again, the majority of respondents believe in heaven according to their answers. In general, the variable is balanced enough for further analysis.

As for missings in the final data set:



```{r}
sum(is.na(usa)) #517
nrow(usa)

f <- function(x) {
  sum(is.na(x))
}

lapply(usa, f)
```



About 20 % of observations are missing values. It is relatively big amoung, however, there are no variables which extremely big number of them. Moreover, for me, it is important to keep all of them for further analysis. Therefore, I will just exclude all NAs:


```{r}
USA <- na.exclude(usa)

summary(USA)
```


As a result, the sample contains 2384 observations. By excluding NAs I lost about 200 of them, however, all variables that are selected are crucial for the analysis. Therefore, having about 90 % of an original sample at the end of all preprocessing steps is a good result.

Before regression analysis, I check relationships between dependent variable (loc_elect) and predictors:


```{r}
t.test(age ~ loc_elect, data = USA)
```


Here, with age, we can see that difference in means of two groups (according to the loc_elect division) is significant: alternative two-tailed hypothesis is accepted and p-value is less than 2.2e-16.


```{r}
table(USA$loc_elect, USA$sex)

chisq.test(USA$loc_elect, USA$sex)

```


As for "sex" variable, here, according to the results of chi-square tests, values are distributed not independently among levels of variables (p-value = 1.263e-15, which is less than 0.05). Therefore, they are not independent.


```{r}
table(USA$loc_elect, USA$relig)

chisq.test(USA$loc_elect, USA$relig)
```


As for riligiosity variable, here, according to the results of chi-square tests, values are distributed not independently among levels of variables (p-value = 2.763e-05, which is less than 0.05). Therefore, they are not independent.


```{r}
table(USA$loc_elect, USA$attend_services)

chisq.test(USA$loc_elect, USA$attend_services)
```


As for services attendance variable, here, according to the results of chi-square tests, values are distributed not independently among levels of variables (p-value = 6.153e-07, which is less than 0.05). Therefore, they are not independent.


```{r}
t.test(god_imp ~ loc_elect, data = USA)
```


Here, with importance of God, we can see that difference in means of two groups (according to the loc_elect division) is significant: alternative two-tailed hypothesis is accepted and p-value is less than 2.2e-16.


```{r}
table(USA$loc_elect, USA$hell)

chisq.test(USA$loc_elect, USA$hell)
```


As for beliefs in hell, it can be concluded that these variables do not depend on each other: p-value = 0.4126, which is higher than 0.05.


```{r}
table(USA$loc_elect, USA$heaven)

chisq.test(USA$loc_elect, USA$heaven)
```


As for beliefs in heaven, it can be concluded that these variables do not depend on each other: p-value = 0.3506, which is higher than 0.05.


```{r}
table(USA$loc_elect, USA$pray_oft)

chisq.test(USA$loc_elect, USA$pray_oft)
```


As for "pray often" variable, here, according to the results of chi-square tests, values are distributed not independently among levels of variables (p-value = 0.0002202, which is less than 0.05). Therefore, they are not independent.




As a conclusion, all selected variables except "hell" and "heaven" will be used as predictors in further analysis.




For building a logit regression model, I use forward selection.
The first model contains only control variables:


```{r}
model1 <- glm(loc_elect ~ age + sex, data = USA, family = "binomial")

summary(model1)

exp(model1$coefficients)
```


First of all, all control predictors have significant effect on the dependent variable (p-values of both predictors less then 0.05).
As for the effect itself, we can more precisely observe it by exponentiating log odds that we have in the original output of logit regression:
- **1.04315671** for age means that with the increase of age on 1 point (in case of age is it 1 year) the odds of taking part in local elections would be 1.04315671 times higher (or + 4 %);
- **1.53728371** for sex means that if person is male, he 1.53728371times higher odds to participate in local elections in comparison to female (or + 50 %). It is really interesting, because I additionally run models for a couple of other countries (just because f curiosity), and in ALL of them sex predictor had insignificant effect. In the US, as we can see, it works.



The next model also includes indicator, which represents how often respondents attend religious services:



```{r}
USA$attend_services <- relevel(USA$attend_services, ref = "Practically never")

model2 <- glm(loc_elect ~ age + sex + attend_services, data = USA, family = "binomial")

summary(model2)

exp(model2$coefficients)
```



First of all, all levels of some levels of service attendance variables have significant effect on the participation in local elections in the USA. Taking into account that reference category here are respondents who attend such services practically never (I change reference, because, in my opinion, it is more interesting to compare all effects with this category), we can make following conclusion:
- people, who attend such services once a month, once a week or more than once a week, have higher odds to participate in local US elections, in comparison with people who practically never attend services (+68 %, + 58 % and + 64 % respectively).


The next model also includes indicator, which represents how God is important for the respondents:


```{r}
model3 <- glm(loc_elect ~ age + sex + attend_services + god_imp, data = USA, family = "binomial")

summary(model3)

exp(model3$coefficients)
```


First of all, all predictors have significant effect on the dependent variable (p-values of all predictors less then 0.05).
Also, we can note a difference in odds for attend_services: for now, for people, who attend such services once a month it is + 110 %, once a week - +116 %, more than once a week - +231 %, in comparison with people who practically never attend services.
Also, now this effect is significant for people that attend services only on Christmas or Easter day: + 60 % odds, compare to respondents from the reference group.


As for the importance of God, this indicator has negative effect on odds in case of local election participation: with the increase of the importance of God, the odds to take part would decrease on about 6 % for each point.

The next model also includes indicator, which represents how respondents identify themselves in case of religiosity:


```{r}
model4 <- glm(loc_elect ~ age + sex + attend_services + god_imp + relig, data = USA, family = "binomial")

summary(model4)

exp(model4$coefficients)
```


Religiosity indicator is insignificant here: all levels of this variables have relatively similar effect on electoral participation at the local level (with some "liberal" pre-assumptions about alpha-level, non-religious respondents would show a difference in participation, compare to the atheists; however, it is not a case here). Therefore, this indicator will be removed from the model. Moreover, it is meaningless to interpret odds results here.

In the next model is included indicator, which represents how often respondents pray:


```{r}
model5 <- glm(loc_elect ~ age + sex + attend_services + god_imp + pray_oft, data = USA, family = "binomial")

summary(model5)

exp(model5$coefficients)
```


"Pray" indicator is insignificant here: p-value of all levels of this variable effect are bigger than 0.05. Therefore, this indicator will be removed from the model. Moreover, it is meaningless to interpret odds results here.

As a interim conclusion, model 3 is a final one. It includes only two predictors: religious services attendance (with positive effect) and importance of God (with negative effect). In general, it can be assumed that services attendance represents a group participation dimension, and God importance is about value side of observed effect. However, interesting that value dimension has negative effect on electoral participation.

As for the hypotheses, one of them is disproved by results:
1. High level of religious participation **is associated** with the high level of electoral participation at the local level.
2. Strong religious beliefs **are not associated** with the high level of electoral participation at the local level.

In order to make sure that we can conclude something about results, it is necessary to check the model fit. I will do with the use of confusion matrix. It represents how many of cases were predicted correctly. For doing so, I will use yardstick package, which allows us to create a detailed report on how model is fitted.
But first of all, I need to create a regular table, in which includes actual value from dataset and predicted by model 3 values:


```{r}
actual_response <- USA$loc_elect

predicted_response <- round(fitted(model3))

(outcomes <- table(predicted_response, actual_response))
```


According to this table, there are 229 false positive and 498 false negative values, which are mistakes of the model.


```{r}
confusion_matrix <- conf_mat(outcomes)

(conf_indicators <- summary(confusion_matrix, event_level = "second"))

autoplot(confusion_matrix, type = "mosaic")
```


Plot represents the same information that is presented above. Here it is clear that number of mistakes is much less than, for example, number of True Negative values, which are predicted correctly and represented by the left upper corner of the plot. However, it is still necessary to check some main cafeterias, such as **accuracy** - how often is the classifier correct ((TP+TN)/total), **sensitivity** - how apt the model is to detecting events in the positive class (TP/ TP + FN), **specificity** - true negative rate (TN / TN + FP):


```{r}
conf_indicators %>% filter(.metric %in% list("accuracy", "spec", "sens"))
```


In general, model is enough well fitted - accuracy = 0.6950503, which means that model predicts correctly in about 70 % of cases.
Also, we have high specificity - model is okay in prediction of true negative values, what we have  already concluded based on the plot.
However, there is a problem with detecting true positive values: their share (sensitivity) in relatively small.
This results of model fitting can be due to the sample: it was a bit unbalanced with the majority of respondents, who do not participate in local elections.

In general, model has rights to exist with such parameters. But it is also important to check another parameter - area under the curve (receiver operating characteristic, which is a graph showing the performance of the arrest model across all penalty thresholds) or **auc**:


```{r, message=FALSE, warning=FALSE}
for_auc = roc(USA$loc_elect, fitted(model3)) 
plot(for_auc, col="red")
auc(for_auc)
```


As we can see, auc = 0.7108. It means that model predicts about 71 % of cases correctly. This is not a big percent, however, this model still have rights to exist.

Because model is fitted relatively goog, we can check marginal effects without any changes in it:


```{r}
m.eff<-allEffects(model3)
m.eff
```


Here we have predicted probabilities to participate in local elections and their changes due to changes in predictors.

For example, if a person is 20 years old, he or she has just about 18 % probability to participate in local US elections. However, with the increase in age we observe the increase in probabilities too: if a person is 50 years old - about 44 % probability to participate, if 80 - even about 75 %. Therefore, age has a strong positive effect of electoral behavior at local level in the USA.

In case of gender we also observe significant difference: if a person is female, she has about 33 % probability to elect at local level. However, if person is male, he has about 42 % probability. This is much higher, but still not really big. Probably, in the US local elections are not popular, which can explain that the majority of respondents in the sample do not participate in them.

As for the religious services attendance, some of the levels of this indicator also have positive effects. For example, (taking into account that effects of such levels as "once a year" and "less than once a year" are insignificant) if a person practically never attends religious services, he or she would have only about 30 % probability to participate in local US elections.
Also, if a person attends religious services only on Christmas or Easter day, he or she would have about 40 % probability to participate in local US elections.
People, who attend such services once a month, once a week or more than once a week, have higher odds to participate in local US election (47 %, 47 % and 49 % respectively).

With "importance of God" variable we observe completely different effect. If God is not important for a person, he or she has about 46 % probability to participate in local US elections. In contrast, if God is very important, a person has about 33 % probability.

The same information is represented by the plots:


```{r, message=FALSE, warning=FALSE}
plot_model(model3, type="pred")
```


Also, it is interesting to check interaction effect between some variables. I hypothesize that we an observe significant effect between gender and age because of their strong influence in previous models:


```{r, warning=FALSE, message=FALSE}
int_eff <- glm(loc_elect ~ age * sex + attend_services + god_imp, data = USA, family = "binomial")

summary(int_eff)

exp(int_eff$coefficients)
```


Of course, this effect is significant. Age effect is stronger fore males: the increase in age for males leads to about 2 % increase in odds to participate in elections.
Also, important to note that sex variable itself becomes insignificant. It is obvious, because this indicator and created interaction shows relatively same things, which is proved by high multicollinearity shown by the variance inflation factor (9.446207 for sex and 11.983007 for interaction).


```{r, message=FALSE, warning=FALSE}
plot_model(int_eff, type = "pred", terms = c("age", "sex")) +
  labs(x = "\nAge\n",
           y = "\nProbability to participate in elections\n") +
  theme_bw()

```


In general, age has positive effect on both genders. However, in case of males it is much stronger and no so linear.

Another method of classification that can be applied here is decision tree. I will create a classification and regression tree (CART), because I am interested not only in estimating numeric values, but also in some classification (in case of sex variable):


```{r, message=FALSE, warning=FALSE}
tree <- rpart(loc_elect ~ age + sex + attend_services + god_imp, data=USA, control=rpart.control(cp=.0001))
prp(tree)
```


For sure, this tree is overfitted. I will prune it, based on the best complexity parameter:


```{r}
best <- tree$cptable[which.min(tree$cptable[,"xerror"]),"CP"]

pruned_tree <- prune(tree, cp=best)

prp(pruned_tree)
```


As a result, we observed that, in general, this classification is based on the all of 4 indicators, presented in the logit regression model: age, gender, religious services attendance and importance of God. The overall logic of classification is the same: higher the age (this variable is a root here) - higher the probability to participate in local US elections;
for males the probability to participate is higher, in comparison with females; more frequent the attendance of services - higher the probability to participate.
higher the importance of God - less the probability to participate in local US elections. However, there is on branch, where the importance of God positively affects electoral participation: for males aged 57 or higher (if importance is not less than 8, prababily is higher, compare to the case if importance is less). As I can see, it is the main difference, compare to the logit regression model: there God importance has only negative effect;
if a person attends religious services more frequently, the probability to participate in local US elections is much higher.
The highest probability to participate have males aged 65 or higher, who frequently attend  religious services.

Another important factor that influence electoral participation at all level, including a local one, in person`s confidence in institutions, which somehow influence the political sphere. However, confidence is a really complicated and latent factor. Even in case of institutions, confidence in different organisations can represent different dimensions of this factor. In such situations, it is necessary to implement **exploratory factor analysis** (EFA) in order to check, which factors are hidden behind variables, which are, from the first look, can be responsible for a single latent factor.


```{r}
efa_data <- usa1 %>%
  select(e069_01, e069_02, e069_04, e069_05, e069_06, e069_07, e069_08, e069_18a, e069_18,
         e069_11, e069_12, e069_13, e069_14, e069_17, e069_20)

```


As it mentioned above, here I analyse confidence and institutions, such as: church, army, press, labour union, police, national parliament, civil services, regional organisations, the EU, national government, political parties, major companies (commercial), environment protection agencies, court and the UN.


```{r}
names(efa_data) <- c("church", "army", "press", "labour_un", "police", "parliament",
                     "civil_serv", "region_org", "EU", "government", "parties",
                     "companies", "environment_pr", "court", "UN")

names(efa_data)
```

```{r}
descr(efa_data)
```


For now, this dataset contains missings that are recoded as negative values. I will recode them as NAs and check, how many of them each variable has.
Moreover, we can see that EU variable does not contain observations at all. This is obvious, because the sample includes respondents only from the USA. Therefore, this variable will be excluded.

```{r}
for (i in names(efa_data)) {
  efa_data[, i] <- ifelse(efa_data[, i] < 0, NA, efa_data[, i])
  print(paste(i, ": ", sum(is.na(efa_data[, i]))))
}

```


All other variables also have some missing values. However, these numbers are insignificant. Therefore, still only EU will be removed from the model.  


```{r}
efa_data <- efa_data %>%
  select(-c(EU))

descr(efa_data) %>%
  select(var, range)

paste("In general", sum(is.na(efa_data)))
```


For now, all variables are positively recoded.

Also, for correct EFA, the majoroty of variables should be correlated with each other. This can be checked by polycoric correlation, which allows us to examine correlation between non-numeric variables (all of selected variables have just 4 levels, therefore, they should be recoded as factors):


```{r, warning=FALSE, message=FALSE}
efa_data1<-as.data.frame(lapply(efa_data,as.factor))

efa_cor <- hetcor(efa_data1) #polychoric correlation
corrplot::corrplot(efa_cor$correlations)
```


As we can see on a plot, the majority of variables are correlated with each other in both positive and negative way.

Next important step is to check, how many factors can we get out of selected variables:


```{r}
fa.parallel(efa_cor$correlations, nrow(efa_data))
```


As we can see, the suggested number of factors is 4. It is also represented on the plot: on x-axis, FA simulated data cross the FA actual data in 4.

In order to implement the EFA, it is necessary to make all variables numeric. Moreover, by pointing out an argument cor="mixed" we implenemt an automatic rotation:


```{r, warning=FALSE, message=FALSE}
efa_data_new <-as.data.frame(lapply(efa_data1,as.numeric))

fa(efa_data_new, 4, cor="mixed")
fa.diagram(fa(efa_data_new, 4, cor="mixed"))
```


Here, we can see information about EFA model, correlations between factors and variables and also plot with factors themselves. As for the metrics, based on which we can include, how well the model works, **the root mean square of residuals** = 0.02, which is less than 0.1 and, therefore, is absolutely okay; **the root mean square error of approximation** index (RMSEA) = 0.082, which is also less than 0.1 and, therefore, is absolutely okay;
**Tucker Lewis Index of factoring reliability** =  0.917 with a threshold of 0.9.


As for these factors, I can assume that the first one MR1 is about confidence in some activists companies and agencies, which, in most cases, operate in more commercial environment.
MR2 is about confidence in political executive power structures of a particular country;
MR3 is about confidence in state forces (army, church, police, court), which, in some cases, do ideological function;
MR4 is about confidence in some international actors.

In my opinion, the executive state forces confidence can play an important role in electoral behavior, even at local level. Therefore, I will test this factor such metric as **Chronbach`s alpha**, which shows us, how consistent this factor is, and, if this indicator will be higher than a threshold of 0.7, I will include this factor in my final regression model:


```{r}
factor <- efa_data_new %>%
  select(police, army, court, church)
```


```{r}
psych::alpha(factor, check.keys = T) #std.alpha = 0.7204883
```


The average Chronbach`s alpha is equal 0.7204883, which is enough to say that this factor is consistent.

For now, I can not include this factor in the model immediately, because I need to create a single variable out of three included. For doing so, I will use an original dataset in order to create new variable. After that, I will add new variable to a USA dataset, which I used before the exclusion of NA. It is necessary to keep all missing values and exclude them only afder adding new variable, because predictors should be the same length:


```{r}
head(usa)

sum(is.na(usa))
```


First of all, I will prepare each variable included in the factor separately. I need to recode all negative values into NAs and also reverse the scales of these indicators, because their maximum values represent minimum confidence and vice versa:


```{r}
unique(usa1$e069_06)

usa1$police <- ifelse(usa1$e069_06 < 0, NA, usa1$e069_06)

usa1$conf_police <- usa1$police * (-1) + 5

table(usa1$police, usa1$conf_police)
  
```

```{r}
unique(usa1$e069_02)

usa1$army <- ifelse(usa1$e069_02 < 0, NA, usa1$e069_02)

usa1$conf_army <- usa1$army * (-1) + 5

table(usa1$army, usa1$conf_army)
```

```{r}
unique(usa1$e069_17)

usa1$court <- ifelse(usa1$e069_17 < 0, NA, usa1$e069_17)

usa1$conf_court <- usa1$court * (-1) + 5

table(usa1$court, usa1$conf_court)
```



```{r}
unique(usa1$e069_01)

usa1$church <- ifelse(usa1$e069_01 < 0, NA, usa1$e069_01)

usa1$conf_church <- usa1$church * (-1) + 5

table(usa1$church, usa1$conf_church)
```


So, all variables were recoded correctly. Therefore, I an able to create one single indicator by getting the average score out of all variables:


```{r}
usa1$polit_conf <- (usa1$conf_army + usa1$conf_police + usa1$conf_court + usa1$conf_church) / 4

descr(usa1$polit_conf)
```


Important that this variable contains only about 7 % of missing values.
Now, I will filter values of this indicator only of the USA:


```{r}
polit_conf_for_usa <- usa1 %>%
  select(polit_conf)
```

```{r}
nrow(usa)
nrow(polit_conf_for_usa)
```


Dataset with previously used variables and with political confidence indicator have the same length, therefore, I can merge them into one dataset.
Before doing so, I will check a distribution of political confidebce variable:


```{r}
descr(polit_conf_for_usa)
```


In the US sample, this variable has only 2.4 % of missing values.


```{r, message=FALSE, warning=FALSE}
polit_conf_for_usa %>%
ggplot(aes(polit_conf)) +
  geom_histogram(bins = 10, fill = "white", colour = 4) +
labs(main = "",
         x = "\nPolitical Confidence\n") +
  theme_bw()
```


The distribution of political confidence variable is just a bit left-skewed, but, in general, is is pretty much similar to a normal one.


```{r}
usa$polit_conf <- polit_conf_for_usa$polit_conf

sum(is.na(usa)) #574
```


Dataset contains only about 25 % of missing values. Exclusion of them will not make further results bias.


```{r}
USA_new <- na.exclude(usa)
```


In order to include political confidence as a predictor, I need to check, does it have any relations with the participation in local elections in the USA: 


```{r}
t.test(polit_conf ~ loc_elect, data = USA_new)
```


According to the results, difference in means of two groups (according to the loc_elect division) is significant: alternative two-tailed hypothesis is accepted and p-value = 1.849e-14


```{r}
USA_new$attend_services <- relevel(USA_new$attend_services, ref = "Practically never")

factor_model <- glm(loc_elect ~ age + sex + attend_services + god_imp + polit_conf, data = USA_new, family = "binomial")

summary(factor_model)

exp(factor_model$coefficients)
```


The effect of variables that were included previously was not changed significantly. The factor of political confidence itself is significant (p-value = 0.000255, which is less than 0.05): with the increase of this factor by one point, we would observe the increase in odds to partici0pate in local US elections on about 37 %.

In general, I can conclude that participation in local US elections can be explained by two major dimensions of religiosity: group participation, which is represented by the religious services attendance and has positive effect; religious beliefs, which are represented by the importance of God and have negative effect.
Also, important to note the role of age and gender: both of them, especially gender (for male in comparison to female), have significant positive effect on participation.
As for the executive power confidence factor, which is derived out of EFA, it also has a significant positive effect of electoral participation at local level. 




**List of Sources**

1. Driskell, R., Embry, E., & Lyon, L. (2008). Faith and politics: The influence of religious beliefs on political participation. Social Science Quarterly, 89(2), 294-314.
2. Smith, L. E., & Walker, L. D. (2013). Belonging, believing, and group behavior: Religiosity and voting in American presidential elections. Political Research Quarterly, 66(2), 399-413.



